---
title: "Introduction to R (Completed Notebook)"
subtitle: "R for Data Science"
author: "LDG"
output: 
  html_notebook:
    number_sections: true
    theme: readable
    highlight: pygments
    toc: true
    toc_float: 
      collapsed: yes
---

# R The calculator

The simplest way to use R is as a calculator. 

For example, the code below calculates $1+1$. Click "run" (the green play button) to see:

```{r calc1}
1+1 # add one and one
```

Notice the part after "#" (inside a code chunk) is not executed because it as a **comment.** Write comments to remind yourself (or rather, your future self, who will read the code later) what the code does.

Here is another example. This code calculates $2^3 + 4 \times 5 + \frac{5}{6}$:

```{r calc2}
2^3 + 4*5 + 5/6 # two to the third plus four times five plus five divided by six
```

## Checkpoint

Calculate $2^2 \times 3^3$:

```{r checkpoint calc1}
2^2 * 3^3
```

Calculate $2/3 + 4/5 + 6/7$:

```{r checkpoint calc2}
2/3 + 4/5 + 6/7
```

Calculate $2^3 + 4*5 + 6^7$:

```{r checkpoint calc3}
2^3 + 4*5 + 6^7
```

## Objects

OK, we can use R to make calculations.

But what if we want to save the output of one calculation and use it in another calculation?

R has memory. What does it remember? Whatever you tell it to! This is called **defining objects**.

Suppose we want to calculate $1+1$ and store it in an object. We have to give it a name. Let's call it "x":

```{r objects1}
x = 1 + 1
```

Notice two things:

1. `=` is the **assignment operator**. It says: "take 1+1 and assign its *output* to `x`.
2. When we run the code, `x = 1 + 1`, nothing spits out. That's because the output is **stored** in `x`.

What happens if we calculate $x+2$

```{r objects2}
x + 2
```

It's the same as calculating $(1+1)+2$. 

Let's create another object called "my_object" equal to $x \times 4$:

```{r objects3}
my_object = x * 4
```

## Vectors

Vectors are objects with multiple values. 

For example, here is a vector with the values $1,2,3,4$ called "data_vector":

```{r vec1}
my_vector = c(1,2,3,4) # create a vector
my_vector # then print it to the screen
```

We can multiply each value by 2:

```{r vec2}
my_vector * 2
```

and we multiply each value by 2 and create a new vector (e.g., "data_new_vector"):

```{r vec3}
data_new_vector = my_vector * 2
```

Vectors can also contain words -- also referred to as "strings" or "characters". 

For instance, here is a vector with the words "hello" and "world":

```{r vec4}
character_vector <- c("hello", "world") # make the vector
character_vector # view it!
```


## Functions

Objects are nouns. Functions are verbs, acting on those nouns. That is programming in R in a nutshell. You have a data set (object). You do things to that data set (e.g., analyze with functions).

All functions in R have parentheses: `function()`. When we created a vector we used the function `c()`.

If we have a vector of data, like `data_new_vector`, we can calculate the sum with `sum()`:

```{r function1}
sum(data_new_vector)
```

Or find the maximum value with `max()`:

```{r function2}
max(data_new_vector)
```

Or find the minimum value with `min()`:

```{r function3}
min(data_new_vector)
```

Or calculate the average or mean with `mean()`:

```{r function4}
mean(data_new_vector)
```


## Checkpoint 

Recall that you create a vector with `c()` and assign it with `=`. For example, the chunk below creates a vector called "v" with values $1,2,3$:

```{r}
v = c(1,2,3)
```

Create a vector called "x" with the values $5, 9, 4, 8, 40, 22$:

```{r checkpoint objects_functions_1}
x <- c(5,9,4,8,40,22)
```

Multiply `x` by 2:

```{r checkpoint objects_functions_2}
x*2
```

Calculate the sum of `x` with the function `sum()`:

```{r checkpoint objects_functions_3}
sum(x)
```

Calculate the average of `x` with `mean()`:

```{r checkpoint objects_functions_4}
mean(x)
```

Calculate the median of `x` with `median()`:

```{r checkpoint objects_functions_5}
median(x)
```

Calculate the standard deviation of `x` with `sd()`:

```{r checkpoint objects_functions_6}
sd(x)
```

# Dataframes

Dataframes -- like what you see in an Excel sheet -- are just vectors chained together. 

Let's load the `tidyverse`:

```{r load tidyverse, message=FALSE}
library(tidyverse)
```

and then load a built-in dataset called "diamonds" with prices of over 50,000 round cut diamonds:

```{r load diamonds}
data("diamonds") # use the function data() to load a built-in dataset
```

You can view the data inside the notebook like so:

```{r view diamonds}
diamonds
```

or you can click on in the "Environment" window pane in the top-right corner of RStudio.

Notice the data contains a variety of data **types**, such as:

* `price`, `carat`, `depth`, `table`: continuous, numerical variables (scale: ratio)
* `color`: categorical variable (scale: nominal, i.e. a classification with no ranking)
* `cut`: categorical variable (scale: ordinal, i.e. you can rank them)

## Summarizing data

Suppose you wanted to calculate the average price of a diamond. You would

1. take the data, THEN
2. summarize it by calculating mean miles per gallon

In R those two steps are:

```{r dplyr1}
diamonds %>% # take the data, THEN
  summarise(mean(price)) # summarize it by calculating average price
```

But not all diamonds are created equal! (According to De Beers.)

Let's group diamonds by their `cut` `class` and then calculate the mean price:

```{r dplyr2, message=FALSE}
diamonds %>% # take the data, THEN
  group_by(cut) %>%  # group the diamonds by cut, THEN
  summarise(mean(price)) # summarize the data by calculating average price by cut
```

### Checkpoint

Group the diamonds by clarity (`clarity`) and then calculate average price:

```{r checkpoint dplyr1, message = FALSE}
diamonds %>% 
  group_by(clarity) %>% 
  summarise(mean(price))
```

Now group diamonds by `cut` and calculate average (`mean()`), median (`median()`) and standard deviation (`sd()`) price. Hint: you only need one call to `summarise()`!

```{r checkpoint dplyr2, message=FALSE}
diamonds %>% 
  group_by(cut) %>% 
  summarise(mean(price), median(price), sd(price))
```

## Visualizing data

One of the most important plots is the **histogram**. This tells you how observations are **distributed**. Understanding distributions is the bedrock of statistical modeling.

Let's plot the **distribution** of prices:

```{r ggplot1}
ggplot(data = diamonds, aes(x = price)) + 
  geom_histogram()
```

The distribution is heavily skewed. Clearly the **average price** is not meaningful! Why? Because if you were to randomly draw from this distribution, you would not get a diamond close to the average.

To see this let's re-plot the distribution but now with a vertical line at the average price:

```{r ggplot1 continued}
ggplot(data = diamonds, aes(x = price)) + 
  geom_histogram() + 
  geom_vline(xintercept = 3932.8, color = "red")
```
Another key plot type is the **scatter plot**. Use this to study **relationships**.

Let's make a scatter plot of `cty` by `hwy`:

Is there a relationship between a diamond's carats (`carat`) and its price?

We might think about a model in which carats determine prices (and not the other way round). That is, prices are a **function** of carats:

$$
\text{price} = f(\text{carats})
$$
This model implies `carat` should be on the x-axis and `price` on the y-axis:

```{r ggplot2}
ggplot(data = diamonds, aes(x = carat, y = price)) + 
  geom_point()
```
Looks so!

### Checkpoint

Plot the distribution of `table` (another diamond attribute)L

```{r checkpoint ggplot 1}
ggplot(data = diamonds, aes(x = table)) + 
  geom_histogram()
```


Replicate the code above but now plot the distribution of `depth` with `geom_density()` so you have kernel density estimate (basically a smooth histogram):

```{r checkpoint ggplot 2}
ggplot(data = diamonds, aes(x = depth)) + 
  geom_density()
```
Based on this plot do you think the average depth of a diamond is meaningful?

## Plots by category

Many times we care about outcomes conditional on some category. 

Just like `dplyr` makes it easy to `ggplot` makes it very easy to plot by a grouping variables. 

There are two approaches. 

### Fill and color

First, create a single panel, and `fill` (histograms, boxplots, kernel densities) or `color` (scatter plots) by the group:

```{r ggplot fill}
ggplot(data = diamonds, aes(x = price, fill = cut)) + 
  geom_histogram()
```

and for a scatterplot:

```{r ggplot color}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + 
  geom_point()
```

### Facetting

But these types of plots are usually hard to read 

Instead you can put each group in its own panel with `facet_wrap()`:

```{r facet histogram}
ggplot(data = diamonds, aes(x = price)) + 
  geom_histogram() + 
  facet_wrap(~cut)
```


```{r facet scatterplot}
ggplot(data = diamonds, aes(x = carat, y = price)) + 
  geom_point() + 
  facet_wrap(~cut) # note the "~" before the grouping variable
```

or you can even make a **grid** of plots by two categories using `facet_grid()`:

```{r facet_grid}
ggplot(data = diamonds, aes(x = price)) + 
  geom_histogram() + 
  facet_grid(color~cut)
```

