---
title: "Memo: Visualize Regression Models"
subtitle: "R for Data Science"
author: "LDG"
output: 
  html_notebook:
    number_sections: true
    theme: readable
    highlight: pygments
    toc: true
    toc_float: 
      collapsed: yes
---

# Dependencies {-}

```{r, message=FALSE}
library(tidyverse)
library(sjPlot) 
```


The package `sjPlot` provides the function `plot_model`. It plots model objects. Learn more here: https://strengejacke.github.io/sjPlot/index.html

# Linear regression

Basic model with `summary()` view:

```{r}
iris %>% 
  lm(formula = Sepal.Length ~ Sepal.Width + Petal.Length + Species, data = .) %>% 
  summary()
```

Plot the coefficients instead:

```{r}
iris %>% 
  lm(formula = Sepal.Length ~ Sepal.Width + Petal.Length + Species, data = .) %>% 
  plot_model(model = .)
```

Spruce it up:

```{r}
iris %>% 
  lm(formula = Sepal.Length ~ Sepal.Width + Petal.Length + Species, data = .) %>% 
  plot_model(model = ., color = "black", sort.est = TRUE, show.values = TRUE) +
  geom_hline(yintercept = 0, color = "gray") + 
  labs(title = "Look at this model", subtitle = "Wow such insight") + 
  theme_sjplot2() + 
  theme_light()
```



# Logitistic Regression

HDMA data (you need to install the package `AER`):

```{r}
data("HMDA", package = "AER")
```

Basic model with `summary()` view:

```{r}
HMDA %>% 
  glm(formula = deny ~ pirat + mhist + chist + hirat + afam, data = ., family = binomial(link = "logit")) %>% 
  summary()
```

Plot the odds-ratios instead:

```{r}
HMDA %>% 
  glm(formula = deny ~ pirat + mhist + chist + hirat + afam, data = ., family = binomial(link = "logit")) %>% 
  plot_model(model = .) 
```

Fine-tune the plot:

```{r}
HMDA %>% 
  glm(formula = deny ~ pirat + mhist + chist + hirat + afam, data = ., family = binomial(link = "logit")) %>% 
  plot_model(model = ., color = "black", sort.est = TRUE) +
  geom_hline(yintercept = 1) + 
  labs(title = "Probability a mortgage application is denied", subtitle = "Boston Housing Data, 1990") + 
  theme_minimal()
```

Plot predicted probabilities of a continuous variable:

```{r}
HMDA %>% 
  glm(formula = deny ~ pirat + mhist + chist + hirat + afam, data = ., family = binomial(link = "logit")) %>% 
  plot_model(model = ., type = "pred", terms="pirat [all]") 
```

